### 2024-03-25 docs/projects/geography/chicago-bikes/index.qmd
Chicago bikes - rerendered

---
<style></style><style>.printedClojure .sourceCode {
  background-color: transparent;
  border-style: none;
}
</style><style>.clay-limit-image-width .clay-image {max-width: 100%}
</style>
<script src="index_files/md-default12.js" type="text/javascript"></script><script src="index_files/md-default13.js" type="text/javascript"></script><script src="index_files/vega14.js" type="text/javascript"></script><script src="index_files/vega15.js" type="text/javascript"></script><script src="index_files/vega16.js" type="text/javascript"></script>

@@

## Workflow

::: {.sourceClojure}
```clojure
(ns index
  (:require [tablecloth.api :as tc]
            [tech.v3.datatype.datetime :as datetime]
            [clojure.string :as str]
            [scicloj.noj.v1.vis.hanami :as hanami]
            [aerial.hanami.templates :as ht]
            [fastmath-clustering.core :as clustering]
            [scicloj.kindly.v4.kind :as kind]))
```
:::

@@

::: {.sourceClojure}
```clojure
(def coord-colnames
  [:start_lat :start_lng :end_lat :end_lng])
```
:::

::: {.sourceClojure}
```clojure
(def processed-trips
  (-> raw-trips
      (tc/add-column :hour (fn [trips]
                             (->> trips
                                  :started_at
                                  (datetime/long-temporal-field :hours))))
      (tc/select-rows (fn [row]
                        (->> coord-colnames
                             (map row)
                             (every? some?))))))
```
:::

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"field":"n","type":"quantitative"},"x":{"field":"hour","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":{"type":"bar","tooltip":true},"width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/0.csv","format":{"type":"csv"}}});</script></div>
```

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300,"data":{"url":"index_files\/1.csv","format":{"type":"csv"}}});</script></div>
```

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.3}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/1.csv","format":{"type":"csv"}}});</script></div>
```

::: {.sourceClojure}
```clojure
(def clustering
  (-> processed-trips
      (tc/select-columns coord-colnames)
      tc/rows
      (clustering/k-means 100)
      (dissoc :data)))
```
:::

@@

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/add-column :cluster (:clustering clustering))
      (tc/group-by [:cluster])
      (tc/without-grouping->
       (tc/order-by (fn [ds]
                      (-> ds :data tc/row-count))
                    :desc)
       (tc/head 5))
      (tc/aggregate {:n tc/row-count
                     :map (fn [trips]
                            [(-> trips
                                 (hanami/plot ht/layer-chart
                                              {:TITLE "Chicago bike trips"
                                               :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                                               :format {:type "topojson"}}
                                                        :mark {:type "geoshape"
                                                               :filled false
                                                               :clip true
                                                               :opacity 0.3}}
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/rule-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :X2 :end_lat
                                                                      :Y2 :end_lng
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :MCOLOR "purple"
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :end_lat
                                                                      :Y :end_lng
                                                                      :MCOLOR "green"
                                                                      :OPACITY 0.01}))]})
                                 as-png)])
                     :hours (fn [trips]
                              [(hour-counts-plot trips)])})
      kind/table))
```
:::

>>>>

---
<style></style><style>.printedClojure .sourceCode {
  background-color: transparent;
  border-style: none;
}
</style><style>.clay-limit-image-width .clay-image {max-width: 100%}
</style>
<script src="index_files/md-default43.js" type="text/javascript"></script><script src="index_files/md-default44.js" type="text/javascript"></script><script src="index_files/vega45.js" type="text/javascript"></script><script src="index_files/vega46.js" type="text/javascript"></script><script src="index_files/vega47.js" type="text/javascript"></script>

@@

## Setup

::: {.sourceClojure}
```clojure
(ns index
  (:require [tablecloth.api :as tc]
            [clojure.math :as math]
            [tech.v3.datatype.datetime :as datetime]
            [tech.v3.datatype.functional :as fun]
            [clojure.string :as str]
            [scicloj.noj.v1.vis.hanami :as hanami]
            [aerial.hanami.templates :as ht]
            [fastmath-clustering.core :as clustering]
            [geo
             [geohash :as geohash]
             [jts :as jts]
             [spatial :as spatial]
             [io :as geoio]
             [crs :as crs]]
            [scicloj.kindly.v4.kind :as kind]
            [scicloj.kindly.v4.api :as kindly])
  (:import (org.locationtech.jts.geom Geometry Point Polygon Coordinate)
           (org.locationtech.jts.geom.prep PreparedGeometry
                                           PreparedLineString
                                           PreparedPolygon
                                           PreparedGeometryFactory)))
```
:::

## Reading the data

@@

## Coordinate conversions

The datasets uses the [WGS84](https://en.wikipedia.org/wiki/World_Geodetic_System)
coordinate system, representing latitude and longitued over the globe.
[EPSG:4326](https://epsg.io/4326)

For metric computations (e.g., k-means clustering),
we need to convert them to a coordinate system
which is locally correct in terms of distances in a region around Chicago:
NAD83 / Illinois East
[EPSG:26971](https://epsg.io/26971)

::: {.clay-dataset}
United States (USA) - Oregon and Washington. [1 3]:

|   Center coordinates |                             Projected bounds |                   WGS84 bounds |
|----------------------|----------------------------------------------|--------------------------------|
| [316133.6 345590.74] | [[216692.35 43649.23] [416809.98 648476.24]] | [[-89.27 37.06] [-87.02 42.5]] |

::: {.sourceClojure}
```clojure
(def crs-transform
  (geo.crs/create-transform (geo.crs/create-crs 4326)
                            (geo.crs/create-crs 26971)))
```
:::

::: {.sourceClojure}
```clojure
(defn wgs84->Chicago
  "Transforming latitude-longitude coordinates
  to local Euclidean coordinates around Seattle."
  [geometry]
  (geo.jts/transform-geom geometry crs-transform))
```
:::

::: {.sourceClojure}
```clojure
(defn lng-lat->local-coords [lng lat]
  (-> (geo.jts/coordinate lng lat)
      geo.jts/point
      wgs84->Chicago
      geo.jts/coord
      ((fn [^Coordinate coord]
         [(.getX coord)
          (.getY coord)]))))
```
:::

Example:

::: {.sourceClojure}
```clojure
(delay
  [(lng-lat->local-coords -89.27 37.06)
   (lng-lat->local-coords -87.02 42.5)])
```
:::

::: {.printedClojure}
```clojure
[[216692.34803057025 44059.50658557552]
 [407952.845339763 648476.2402163395]]

```
:::

## Preprocessing

::: {.sourceClojure}
```clojure
(def processed-trips
  (-> raw-trips
      ;; Compute starting hours:
      (tc/add-column :hour (fn [trips]
                             (->> trips
                                  :started_at
                                  (datetime/long-temporal-field :hours))))
      ;; Make sure the latitude and longitude are nonmissing:
      (tc/select-rows (fn [row]
                        (->> [:start_lat :start_lng :end_lat :end_lng]
                             (map row)
                             (every? some?))))
      ;; Add local Chicago coordinates:
      (tc/map-columns :start-local-coords [:start_lng :start_lat] lng-lat->local-coords)
      (tc/map-columns :end-local-coords [:end_lng :end_lat] lng-lat->local-coords)
      (tc/map-columns :start-local-x [:start-local-coords] first)
      (tc/map-columns :start-local-y [:start-local-coords] second)
      (tc/map-columns :end-local-x [:end-local-coords] first)
      (tc/map-columns :end-local-y [:end-local-coords] second)))
```
:::

## Comparing Eucledian (L2) distances in global and local coordinates

::: {.sourceClojure}
```clojure
(-> processed-trips
    (tc/random 1000 {:seed 1})
    (tc/add-column :local-L2 #(fun/sqrt
                               (fun/+ (fun/sq (fun/- (:start-local-x %)
                                                     (:end-local-x %)))
                                      (fun/sq (fun/- (:start-local-y %)
                                                     (:end-local-y %))))))
    (tc/add-column :global-L2 #(fun/sqrt
                                (fun/+ (fun/sq (fun/- (:start_lat %)
                                                      (:end_lat %)))
                                       (fun/sq (fun/- (:start_lng %)
                                                      (:end_lng %))))))
    (hanami/plot ht/point-chart
                 {:X :local-L2
                  :Y :global-L2}))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"field":"global-L2","type":"quantitative"},"x":{"field":"local-L2","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":{"type":"circle","tooltip":true},"width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/0.csv","format":{"type":"csv"}}});</script></div>
```

The `local-L2` quantity, compted from the local Chicago coordinates,
is a decent approximation of actual distance.
We see that the `global-L2` quantity, computed accordingly
from latitude and longitude, is not directly related to
this quantity and might be misleading for metric tasks
such as clustering.

## Basic analysis and visualization

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"field":"n","type":"quantitative"},"x":{"field":"hour","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":{"type":"bar","tooltip":true},"width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/1.csv","format":{"type":"csv"}}});</script></div>
```

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/random 1000 {:seed 1})
      (hanami/plot ht/rule-chart
                   {:X :start-local-x
                    :Y :start-local-y
                    :X2 :end-local-x
                    :Y2 :end-local-y
                    :XSCALE {:zero false}
                    :YSCALE {:zero false}})))
```
:::

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"encoding":{"y":{"scale":{"zero":false},"field":"start-local-y","type":"quantitative"},"x":{"scale":{"zero":false},"field":"start-local-x","type":"quantitative"},"y2":{"scale":{"zero":false},"field":"end-local-y","type":"quantitative"},"x2":{"scale":{"zero":false},"field":"end-local-x","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

@@

```{=html}
<div><script>vegaEmbed(document.currentScript.parentElement, {"usermeta":{"embedOptions":{"renderer":"svg"}},"width":400,"background":"floralwhite","title":{"text":"Chicago bike trips"},"layer":[{"data":{"url":"notebooks\/data\/chicago.geojson","format":{"type":"topojson"}},"mark":{"type":"geoshape","filled":false,"clip":true,"opacity":0.3}},{"encoding":{"longitude":{"field":"start_lng","type":"quantitative"},"opacity":{"value":0.1},"latitude":{"field":"start_lat","type":"quantitative"},"longitude2":{"field":"end_lng","type":"quantitative"},"latitude2":{"field":"end_lat","type":"quantitative"}},"usermeta":{"embedOptions":{"renderer":"svg"}},"mark":"rule","width":400,"background":"floralwhite","projection":{"type":"mercator"},"height":300}],"height":300,"data":{"url":"index_files\/2.csv","format":{"type":"csv"}}});</script></div>
```

## Clustering

::: {.sourceClojure}
```clojure
(def clustering
  (-> processed-trips
      (tc/select-columns [:start-local-x
                          :start-local-y
                          :end-local-x
                          :end-local-y])
      tc/rows
      (clustering/k-means 100)
      (dissoc :data)))
```
:::

@@

::: {.sourceClojure}
```clojure
(delay
  (-> processed-trips
      (tc/add-column :cluster (:clustering clustering))
      (tc/group-by [:cluster])
      (tc/without-grouping->
       (tc/order-by (fn [ds]
                      (-> ds :data tc/row-count))
                    :desc)
       (tc/head 20))
      (tc/aggregate {:n tc/row-count
                     :map (fn [trips]
                            [(-> trips
                                 (hanami/plot ht/layer-chart
                                              {:TITLE "Chicago bike trips"
                                               :LAYER [{:data {:url "notebooks/data/chicago.geojson"
                                                               :format {:type "topojson"}}
                                                        :mark {:type "geoshape"
                                                               :filled false
                                                               :clip true
                                                               :opacity 0.3}}
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/rule-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :X2 :end_lat
                                                                      :Y2 :end_lng
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :start_lat
                                                                      :Y :start_lng
                                                                      :MCOLOR "purple"
                                                                      :OPACITY 0.01}))
                                                       (as-geo
                                                        (hanami/plot nil
                                                                     ht/point-chart
                                                                     {:X :end_lat
                                                                      :Y :end_lng
                                                                      :MCOLOR "green"
                                                                      :OPACITY 0.01}))]})
                                 as-png)])
                     :hours (fn [trips]
                              [(hour-counts-plot trips)])})
      kind/table))
```
:::
